Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-10 doc/ApTranscriptome_manuscript.Rmd
added site climate info and species CTmax to Intro

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* (mean annual temperature at collection site = ...) experiences significantly warmer and less seasonally variable temperatures than *A. picea* (mean annual temperature from collection site = ; [@hijmans2005]), with correspondingly higher minimal and maximal critical thermal limits that allow us to investigate adaptation to both lower and upper thermal extremes (...CTmin and CTmax from HF/DF...) [@diamond2012].

>>>>

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* (mean annual temperature (MAT) at collection site = 14.6°C, seasonality = 7,678) experiences significantly warmer and less seasonally variable temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (40.75°C for *A. carolinensis* in Raleigh, North Carolina compared to 38°C for *A. picea* at Harvard Forest, Massachusetts [@diamond2012]). Critical minimum temperatures have been shown to co-vary with critical maximum temperatures between and within *A. picea* and *A. rudis* [@warren2013], allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

```{r climate, echo=FALSE, message=FALSE, warning=FALSE}
library(raster)
w <- getData('worldclim', var='bio', res=10)
#names(w)

# Plot first bioclim dimension (MAT) to check download worked
#plot(w, 1, col=colorRampPalette(c("blue","green","red"))(255))

# sites
sites <- data.frame(site = c("Ac", "Ap"), lon=c(-78.874, -72.702), lat= c(36.037, 44.508))

# extract bioclimatic variables from worldclim data. Note bioclim temperature data is multiplied by 10 (no decimal), so transform by dividing by 10.

dbio <- extract(w, sites[,c("lon","lat")])
# bio1 = MAT
# bio4 = Temperature Seasonality (standard deviation *100)

dbio.vars <- cbind(sites, dbio[,1], dbio[,4])
colnames(dbio.vars)[which(colnames(dbio.vars) == "dbio[, 1]")] <- "MATx10"
dbio.vars$MAT <- dbio.vars$MATx10/10
colnames(dbio.vars)[which(colnames(dbio.vars) == "dbio[, 4]")] <- "Seasonality"

dbio.vars
```
